---
title: "In-Class Exercise 4: An Exploratory Data Analysis Checklist"
subtitle: "Roger Peng's checklist, applied to Freedom on the Move"
author: "Test Student"
date: "October 7, 2026"
format:
  html:
    theme: cosmo
    highlight-style: dracula
    code-overflow: wrap
    css: assets/styles.css
    df-print: paged
    toc: true
    toc-depth: 2
    embed-resources: true
---

::: {.callout-note}
## What is this exercise?

In Exercise 3 you looked closely at how Freedom on the Move (FOTM) records *people*: names, kin, racial descriptions. Today you step back and look at the database as a whole, using a checklist written by the statistician Roger D. Peng for the first look at any new dataset.

This exercise is **adapted from Roger D. Peng, [*Exploratory Data Analysis with R*](https://bookdown.org/rdpeng/exdata/), Chapter 4, "[Exploratory Data Analysis Checklist](https://bookdown.org/rdpeng/exdata/exploratory-data-analysis-checklist.html)."** This chapter, as well as the following one, are listed as your reading for today's class. Peng works through his checklist with a dataset of hourly air pollution readings. Each section below opens with the step from his chapter, summarized, and then applies it to FOTM.

As before, you're graded on your documented effort to understand what you're doing, not on mastery. Start in class (pairs or groups of 3 are welcome), finish at home, and commit your own work to GitHub.
:::

::: {.callout-warning}
## Note on harmful language

As in Exercise 3, the ads you'll read describe people in the language of those who enslaved them. We read these words as historical evidence, and we make a commitment as a class to work with these records mindfully and aware of the violence in them. In other words: we will avoid repeating harmful language where it does not contribute to an interpretation or a historical argument.
:::

## Setup

You need the FOTM tables in `data/FOTM/`, as in Exercise 3. This chunk loads the tidyverse, makes a folder for today's output, and defines `read_fotm()`, the same helper you used in Exercise 3:

```{r}
#| message: false
library(tidyverse)
dir.create("output/exercise4", recursive = TRUE, showWarnings = FALSE)

read_fotm <- function(table) {
  file <- str_glue("data/FOTM/{table}.csv")
  suppressWarnings(                                   # the trailing-comma warning from Exercise 3
    read_csv(file,
             col_types = cols(.default = col_character()),   # read everything as text, change types later
             name_repair = "unique_quiet")
  ) %>%
    select(-starts_with("...")) %>%                   # drop the empty column the trailing comma creates
    rename_with(~ .x %>% str_remove("^\\S+ ") %>% str_to_lower())   # "ADVERTISEMENT ID" becomes "id"
}
```

------------------------------------------------------------------------

## 1. Formulate your question

> **Peng:** "A sharp question or hypothesis can serve as a dimension reduction tool that can eliminate variables that are not immediately relevant to the question." It's also worth asking early whether you have the right data to answer it. ([Peng, "Formulate your question"](https://bookdown.org/rdpeng/exdata/exploratory-data-analysis-checklist.html#formulate-your-question))

Peng starts from a broad question ("Are air pollution levels higher on the east coast than on the west coast?") and sharpens it until it names one measure and one unit. We'll do the same. A broad question:

> Where did people escape from slavery?

FOTM can't answer that directly: it holds *advertisements*, not escapes. 


(@) Write your own sharp question about FOTM's advertisements: one that names exactly what you'd count or measure, and for what unit (an ad, a newspaper, a state, a decade...).

> how many ads were placed by jailers, compared with enslavers, in each decade from 1800 to 1860?

Save your question for a possible future investigation. For the purpose of learning Peng's checklist, we'll pursue a question this database can actually answer:

> In which states were the advertisements in FOTM published?

------------------------------------------------------------------------

## 2. Read in your data

> **Peng:** Read the data in, and note anything you had to do to make it readable. ([Peng, "Read in your data"](https://bookdown.org/rdpeng/exdata/exploratory-data-analysis-checklist.html#read-in-your-data))

Our question needs three tables: the ads, the newspapers they appeared in, and where those newspapers were published.

```{r}
ads        <- read_fotm("advertisement")   # one row per ad
newspapers <- read_fotm("newspaper")       # the newspaper each ad appeared in
locations  <- read_fotm("location")        # where each newspaper was published
```

Join them, the same way you joined tables in Exercises 1 and 3. Count the rows before and after. A join that has more rows than the original table is a warning sign.

```{r}
nrow(ads)
```

```{r}
ads_place <- ads %>%
  left_join(newspapers %>% select(newspaper_id = id, newspaper = name, location_id),
            by = "newspaper_id") %>%
  left_join(locations %>% select(location_id = id, city, state),
            by = "location_id")

nrow(ads_place)   # should be the same as nrow(ads)
```

------------------------------------------------------------------------

## 3. Check the packaging

> **Peng:** Before opening the wrapped "present," shake the box: how many rows and columns are there? And do they match what you expected? ([Peng, "Check the packaging"](https://bookdown.org/rdpeng/exdata/exploratory-data-analysis-checklist.html#check-the-packaging))

```{r}
nrow(ads)
ncol(ads)

nrow(newspapers)
n_distinct(newspapers$name)   # how many different newspaper titles?
```

Titles aren't the whole story: two papers can share a name, and one paper can move. So count the different newspapers by title **and** place of publication:

```{r}
ads_place %>%
  distinct(newspaper, city, state) %>%
  nrow()                               
```

(@) What did the `distinct()` function do in the chunk above?
Hint: take a peak at the documentation for the function with `?distinct`.

> `distinct(newspaper, city, state)` keeps one row for each unique combination of title, city, and state and drops the duplicates. It returns only those three columns. Piping the result into `nrow()` counts the combinations: 261 different newspapers, by title and place.

(@) The `newspaper` table has 41,551 rows, but notice many different newspapers (unique pair of title + place) there are. So what is the **unit of analysis** in the `newspaper` table? In other words, what is one row? 

> 41,551 rows describe only 261 newspapers (title + place) used by the ads. The intended unit of analysis is technically one newspaper title in its place of publication. 
>
> But in most cases, FOTM created a new newspaper record with its own location record for every advertisement transcribed into the database. 41,303 newspaper IDs are used by exactly one ad. The Daily National Intelligencer has 24,311 rows, one for each of its 24,311 ads. A few titles reuse one record across thousands of advertisements.
>
> Ultimately, the table's ID does not define its unit of analysis. What the rows describe does. To count newspapers, we must group by title and place instead of counting rows of `newspaers` alone.

(@) Then find a title that appears in two places: did the paper move, are these two different papers, or is it an error?

::: {.callout-tip}
## Tip:
To answer the question, you need two things for each title: **where** it was published, and **when**. Build your chunk in three steps:

1. **One row per title and place, with the evidence.** Start from `ads_place`, group by title and place, and summarize how many ads each one has and when they were published:

```r
   ads_place %>%
     group_by(newspaper, city, state) %>%
     summarize(ads   = n(),
               first = min(publication_date),
               last  = max(publication_date),
               .groups = "drop")
```
2. **Count places per title.** Pipe that into `add_count(_______)` so that each row shows how many places its title appears in.

3. **Keep only titles with more than one place:** `filter(___ > ___)`.

**How to read the result:**

- **Dates follow one another** (one place stops, the other starts): the paper probably **moved**.
- **Dates overlap, or are years apart:** probably **two different papers** with the same name.
- **Same dates in both places, or a single stray ad:** probably a **data-entry error**.
:::


```{r}
 ads_place %>%
  group_by(newspaper, city, state) %>%
  summarize(ads   = n(),
            first = min(publication_date),
            last  = max(publication_date),
            .groups = "drop") %>% 
  add_count(newspaper) %>% 
  filter(n > 1)
```
Write your findings here:

> Seven titles appear in two places (14 rows):
>
> - *Daily National Intelligencer:* Washington, NY (1) and Wasshington, DC (24,310). Possibly a data error, as one ad was placed in New York.
> - *Daily Southern Citizen:* Edenton (1) and Greensboro (6), both 25 June 1864. Likely an error, since the same issue (published on the same date) was given two different locations.
> - *Weekly Courier:* Similarly, probably an error: same dates in 1860, two different locations.
> - *State Gazette of North Carolina:* The date ranges for the different locations follow one another, which suggests the title possibly moved.
> - *People's Press:* The two records are 15 years apart, so it is likely that these are two different newspapers.
> - *North Carolina. Journal:* The range of publication date overlaps for both locations. So these are either two different newspapers or the same newspaper was published in different places at different moments.
> - *Newbern Spectator:* One single ad in Raleigh in 1865, possibly an error.
>
> These readings come from the data alone. We would need to crossreference with outside sources to check whether these explanations are sound.

------------------------------------------------------------------------

## 4. Run `str()`

> **Peng:** `str()` shows the class of every column. Make sure numbers are numbers, dates are dates, and text is text. ([Peng, "Run str()"](https://bookdown.org/rdpeng/exdata/exploratory-data-analysis-checklist.html#run-str))

```{r}
str(ads_place)
```

Every column is text, because `read_fotm()` reads everything as text on purpose. To work with time we need a real date. `year()` (from the lubridate package, part of the tidyverse) pulls out the year:

```{r}
ads_place <- ads_place %>%
  mutate(date = as.Date(publication_date),
         year = year(date))

sum(is.na(ads_place$date))   
# how many dates failed to convert?
```

------------------------------------------------------------------------

## 5. Look at the top and the bottom of your data

> **Peng:** Look at the first and last rows. Problems often hide at the end of a file, and for data over time, check that the first and last dates are what you expect. ([Peng, "Look at the top and the bottom of your data"](https://bookdown.org/rdpeng/exdata/exploratory-data-analysis-checklist.html#look-at-the-top-and-the-bottom-of-your-data))

Our data are over time, so sort by date first:

```{r}
ads_place %>%
  arrange(date) %>%
  select(date, newspaper, city, state) %>%
  head()

ads_place %>%
  arrange(date) %>%
  select(date, newspaper, city, state) %>%
  tail()
```

(@) Discuss with your pair or group: what are the earliest and latest dates? Which one surprises you, and why?

> Earliest: 26 June 1704, *The Boston News-Letter* (Boston, MA). 
> Latest: 25 April 1894, *Weekly Raleigh Register* (Raleigh, NC).
>
> The latest date stands out: it is 30 years after slavery ended. All the last rows (all North Carolina) are also after the end of the Civil War in 1865.

------------------------------------------------------------------------

## 6. Check your "n"s

> **Peng:** Count things, and compare the counts with *landmarks*: numbers you know should be true. If you expect 1,000 people and find 20, something's wrong. ([Peng, "Check your 'n's"](https://bookdown.org/rdpeng/exdata/exploratory-data-analysis-checklist.html#check-your-ns))

A landmark for us: every ad in FOTM is a scan of a newspaper page. If FOTM is a database of *transcribed* ads, nearly every row should have a transcription. Does it? (`{undefined value}` is one of FOTM's ways of saying "nothing here," as you saw in Exercise 3.)

```{r}
ads_place <- ads_place %>%
  mutate(has_text = !is.na(transcription) & transcription != "{undefined value}")

ads_place %>% count(has_text)
```

Plot the ads by decade, colored by whether they've been transcribed. `floor(year / 10) * 10` turns 1823 into 1820:

```{r}
ads_place %>%
  mutate(decade = floor(year / 10) * 10) %>%
  count(decade, has_text) %>%
  ggplot(aes(x = decade, y = n, fill = has_text)) +
  geom_col() +
  labs(title = "FOTM advertisements by decade of publication",
       x = "Decade", y = "Advertisements", fill = "Transcribed?")
```

(@) What share of the ads have a transcription? Which decades have the most untranscribed ads? Without the color, what would a reader conclude from the height of the 1810s and 1820s bars?

> Only 27,131 of 89,601 ads (about 30%) have a transcription; 62,470 don't.
>
> The most untranscribed ads are in the 1820s, the 1810s, and the 1800s.
>
> Without the color, a reader would conclude that runaway advertising (or escape itself) peaked in the 1810s–1820s. In fact the spike mostly reflects FOTM having catalogued a long run of one newspaper, the *Daily National Intelligencer* (1813–1831), whose ads are mostly not yet transcribed. The bars show the archive's composition instead of pointing to a real historical trend.

------------------------------------------------------------------------

## 7. Validate with at least one external data source

> **Peng:** Check your data against something *outside* the dataset. Even a single number will do: Peng compares his ozone readings with the national air-quality standard to see if they're the right order of magnitude. ([Peng, "Validate with at least one external data source"](https://bookdown.org/rdpeng/exdata/exploratory-data-analysis-checklist.html#validate-with-at-least-one-external-data-source))

Two historical landmarks:

- *The Boston News-Letter*, the first continuously published newspaper in British North America, began in April **1704**.
- The Thirteenth Amendment abolished slavery in the United States in December **1865**.

Your earliest ad should be no earlier than 1704, and runaway-slave ads should stop after 1865. Check the first, then read the ads published after 1865:

```{r}
ads_place %>% filter(year == min(year)) %>% select(date, newspaper)

after_1865 <- ads_place %>%
  filter(year > 1865) %>%
  arrange(date)

after_1865 %>% count(newspaper, state)

after_1865 %>% pull(transcription) %>% cat(sep = "\n\n----------\n\n")
```

(@) Does the earliest ad agree with the first landmark? For the ads after 1865: who is running away, from whom, and who "bound" them? What do these ads tell you about the end of slavery, and why might FOTM's editors have included them? Is any of them probably a data-entry error?

> Yes, the earliest ad is from June 26, 1704, two months after *The Boston News-Letter* began publishing (24 April 1704).
>
> There are 15 ads recorded after 1865, all from North Carolina. Twelve are from the *Greensborough Patriot* (1866–1868), two from the *Weekly Pioneer* (Asheville, 1872 and 1874), and one from the *Weekly Raleigh Register* (1894).
>
> Most of the *Patriot* ads concern Black children and teenagers. They were "bound" or "indentured" to white employers by the Guilford County Court or by the Freedmen's Bureau's Assistant Superintendent. The rewards are one, five, or ten cents: the point was to warn others not to "harbor or employ" them, not to recover their value.
> These ads suggest that coerced labor continued after Emancipation through the apprenticeship ("binding out") of Black children, which North Carolina and other Southern states used after 1865. Emancipation was not a clean break in the sources. FOTM editors might have decided to include these ads because they belong to the same advertising genre as other common ads. They start with "Ranaway," included a reward, a description, and a warning not to harbor. One transcription includes a volunteer's note, "Not an enslaved person. All entries will be left blank so as not to be mistaken in the system." So the editors kept the ad but declined to mark the person up as enslaved.
>
> On the other hand, a few ads seem to be outliers. An ad from 1872 listed a sheriff's $25 reward for capturing a man named Andrew Rucker (law enforcement, not slavery). Similarly, an 1874 ad searched for two white boys, one a bound apprentice.
>
> **Data-entry error:** the "1894" ad describes an escape in July 1847 and is signed "April 16, 1849." Its publication date was entered wrongly, which speaks to the challenges of crowdsourcing for data entry and transcription.


------------------------------------------------------------------------

## 8. Try the easy solution first

> **Peng:** Give the simplest possible answer to your question first, without worrying yet about whether it's right. Then question it. ([Peng, "Try the easy solution first"](https://bookdown.org/rdpeng/exdata/exploratory-data-analysis-checklist.html#try-the-easy-solution-first))

The easy answer to "In which states were the ads published?": count the ads by state.

```{r}
ads_by_state <- ads_place %>%
  count(state, name = "ads", sort = TRUE)

ads_by_state

ggplot(ads_by_state, aes(x = ads, y = reorder(state, ads))) +
  geom_col() +
  labs(title = "FOTM advertisements by state of publication",
       x = "Advertisements", y = NULL)
```

(@) Which state comes first? Does that match what you know about where slavery was concentrated in the United States?

> The District of Columbia, with 26,929 ads, ahead of South Carolina (13,930), Virginia (12,231), and Louisiana (11,790).
>
> It does not match where slavery was concentrated. In 1860, for example, D.C. had about 3,200 enslaved people, while Virginia, Georgia, Mississippi, Alabama, and South Carolina each had more than 400,000.


------------------------------------------------------------------------

## 9. Challenge your solution

> **Peng:** The easy answer should never "hold the day." Look for ways to challenge it, especially when it matches what you expected. ([Peng, "Challenge your solution"](https://bookdown.org/rdpeng/exdata/exploratory-data-analysis-checklist.html#challenge-your-solution))

### Challenge 1: Which newspapers?

A state's count is the sum of its newspapers. Which newspapers make up the top state?

```{r}
top_state <- ads_by_state$state[1]

ads_place %>%
  filter(state == top_state) %>%
  count(newspaper, has_text, sort = TRUE)
```

### Challenge 2: Count only the ads someone has read

Redo the easy solution, counting only transcribed ads:

```{r}
ads_place %>%
  filter(has_text) %>%
  count(state, name = "transcribed_ads", sort = TRUE)
```

(@) How much of the top state's lead comes from one newspaper? How many of that paper's ads have been transcribed? Which state comes first when you count only transcribed ads?

> The *Daily National Intelligencer* accounts for 24,310 of DC's 26,929 ads (about 90%). Only 168 of them were transcribed. The *Washington DC National Intelligencer*, with 2,619 ads and 1,908 transcribed, is very likely the same paper under a second name in the database.
>
> When counting only transcribed ads, Louisiana comes first (8,041, mostly from *The Daily Picayune*), then North Carolina (5,186) and South Carolina (3,414). DC drops to sixth place (2,076, tied with Alabama).The ranking depends on what you count, and the "easy" answer mostly measured one newspaper's untranscribed ads.

### Challenge 3: Compare with the enslaved population

A second external source: the number of enslaved people in each slave state in 1860, from the U.S. Census. It's only a rough yardstick (the ads span 160 years, the census is one year), but it shows the order of magnitude we should expect.

```{r}
census_1860 <- tribble(
  ~state,  ~state_name,            ~enslaved_1860,
  "US-AL", "Alabama",               435080,
  "US-AR", "Arkansas",              111115,
  "US-DE", "Delaware",                1798,
  "US-DC", "District of Columbia",    3185,
  "US-FL", "Florida",                61745,
  "US-GA", "Georgia",               462198,
  "US-KY", "Kentucky",              225483,
  "US-LA", "Louisiana",             331726,
  "US-MD", "Maryland",               87189,
  "US-MS", "Mississippi",           436631,
  "US-MO", "Missouri",              114931,
  "US-NC", "North Carolina",        331059,
  "US-SC", "South Carolina",        402407,
  "US-TN", "Tennessee",             275719,
  "US-TX", "Texas",                 182566,
  "US-VA", "Virginia",              490865
)

census_1860 %>%
  left_join(ads_by_state, by = "state") %>%
  mutate(ads = replace_na(ads, 0),                      # no ads at all becomes 0, not NA
         ads_per_1000_enslaved = round(ads / enslaved_1860 * 1000, 1)) %>%
  arrange(desc(ads_per_1000_enslaved))
```

(@) Which states have the most ads per 1,000 enslaved people, and which have none at all? What does that suggest about how FOTM chose (or was able to find) its newspapers?

> The District of Columbia had the most ads per 1,000 enslaved people (about 8,455), far ahead of Maryland (85), Louisiana (35.5), and South Carolina (34.6).
> 
> Arkansas, Mississippi, and Missouri had none at all. Kentucky has nearly none (22 ads). Mississippi had 436,631 enslaved people in 1860 and has no ads in FOTM.
> 
> This suggests that the counts reflect which newspapers FOTM could find, digitize, and catalogue (surviving runs, partner collections, transcription progress), not where enslaved people lived or escaped. A state's absence from FOTM is a gap in the archive, not necessarily in history.

(@) Last challenge, from Exercise 3: the George, Celia, and Sarah ad ran 20 times in the *Charleston Mercury*, and each printing is a separate row in `ads`. How would reprints affect the counts above? How might you find them? (You don't need to write the code.)

> Reprints inflate the counts. One escape advertised 20 times becomes 20 rows. Papers and states with long runs of reprinted ads, or more complete digitized runs, look bigger.
> To find reprints, searching for a match on newspaper + advertiser + the opening words of the text or comparing the people and dates in the escape records are possible paths.

------------------------------------------------------------------------

## 10. Follow up

> **Peng:** At the end, ask three questions. Do you have the right data? Do you need other data? Do you have the right question? ([Peng, "Follow up questions"](https://bookdown.org/rdpeng/exdata/exploratory-data-analysis-checklist.html#follow-up-questions))

(@) Answer Peng's three questions for our question ("In which states were the advertisements in FOTM published?"). Then return to the broad question we set aside in step 1, "Where did people escape from slavery?" What would you need to answer it, and what can FOTM honestly tell you about it?

> **Right data?** Only partly. FOTM tells us where the ads included in the database were published. The counts mostly measure FOTM's coverage: which newspapers were digitized and which ads have been transcribed. One DC paper dominates.
> **Other data needed?** A directory of newspapers that existed (e.g., the Library of Congress's Chronicling America directory), could help confirming what's missing. The census can offer a cross-reference point for the enslaved population. Information on FOTM's digitization partners and transcription progress would put the database structure in context.
> **Right question?** A better one: "How does FOTM's coverage of transcribed, de-duplicated ads vary by state and decade?"
> **Where did people escape from slavery?** To answer this question, we need records of *escapes*, not ads. Within FOTM, that means the escape events and the places marked "ran from" in `runaway_location`, but only for the ads volunteers have tagged, with reprints removed. Other sources would help too: jail registers, court records, Underground Railroad records, the accounts of people who escaped, etc.

(@) Rewrite the question you wrote in step 1, using what you learned in this checklist.

> "Among the transcribed and tagged ads in FOTM, de-duplicated for reprints, how did the share of ads placed by jailers change between 1800 and 1860, and in which newspapers?"

------------------------------------------------------------------------

## Save, Commit, Push

The counts you made are small summaries of FOTM, so they can be committed. FOTM's own files stay in `data/FOTM/`.

```{r}
write_csv(ads_by_state, "output/exercise4/ads_by_state.csv")
```

Commit this file and `output/exercise4/` with a message like "Exercise 4: EDA checklist on FOTM," then **Push origin**. As always, if anything from `data/FOTM/` shows up in GitHub Desktop's Changes list, stop and ask me.

::: {.callout-tip}
## Sources

- Roger D. Peng, *Exploratory Data Analysis with R* (Leanpub, 2015–2020), Chapter 4, "Exploratory Data Analysis Checklist," [bookdown.org/rdpeng/exdata](https://bookdown.org/rdpeng/exdata/exploratory-data-analysis-checklist.html). The structure of this exercise and the summarized steps are his.
- *Freedom on the Move: A Database of Fugitives from American Slavery*, Cornell University, [freedomonthemove.org](https://freedomonthemove.org). Full database download used in this course: September 2026.
- Enslaved population by state, 1860 U.S. Census, as tabulated in "[Population by Color and Status: U.S. Slave States, 1860](https://terpconnect.umd.edu/~sfmiller/population%20statistics,%201860.htm)."
:::
